Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2020-10-26 paper/main.tex
more comments and fixes

\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
\usepackage{booktabs} % For formal tables
\usepackage{enumitem}
\usepackage{afterpage,tabularx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage[square,numbers,sort]{natbib}
% Fixes for using natbib and llncs
\makeatletter
% required for natbib to have "References" printed and as section*, not chapter*
\renewcommand\bibsection%
{\section*{\refname\@mkboth{\MakeUppercase{\refname}}{\MakeUppercase{\refname}}}}
\renewcommand\@biblabel[1]{#1.}
\def\bibfont{\small}
\makeatother

@@

Distances in permutation space can be measured in various ways.  Kendall's-$\tau$ distance counts the number of items ranked in different order in both permutations:
%
\begin{equation}
\begin{split}
d(\sigma _{1},\sigma_{2})=\left\lvert\Bigl\{(i,j):i<j \land \phantom{}\Bigl(\right.&\bigl(\sigma_{1}(i) < \sigma_{1}(j)\land \sigma_{2}(i)>\sigma_{2}(j)\bigr)\lor \phantom{} \\
 &\bigl(\sigma_{1}(i)>\sigma_{1}(j)\land \sigma_{2}(i)<\sigma_{2}(j)\bigr)\Bigr)\Bigr\}\Bigr\rvert\\.
 \end{split}
 \label{eq:kendall_decomp}
\end{equation}

An equivalent\MANUEL{X Alternative or equivalent? That is, it is the same numbers or it will produce a different result?} definition for the Kendall's-$\tau$ distance counts the number of adjacent swaps that have to be made in $\sigma_1^{-1}$ to obtain $\sigma_2^{-1}$ and, therefore, it is sometimes called \emph{swap distance}~\citep{ZaeStoBar2014:ppsn}.

@@

\subsection{Distributions over permutations}
Distributions over permutations~\cite{critchlow91} are functions that assign a probability value  $\Prob(\sigma)\in[0,1]$ to each permutation $\sigma \in S_n$. One of the most popular distributions is the Mallows Model (MM), which is considered as an analogous to the Gaussian distribution for permutations. The MM defines the probability of each permutation $\sigma\in S_n$ as follows:\MANUEL{X Could you check this? I wanted to introduce the notation $MM(\sigma_0, \theta)$ that we use later}
%
\begin{equation}\label{eq:MM}
\Prob(\sigma \mid \sigma_0, \theta )= \frac{\exp(-\theta d(\sigma, \sigma_0))}{\psi} \sim MM(\sigma_0, \theta)
\end{equation}
%
with two parameters, $\theta$ and $\sigma_0$. Permutation $\sigma_0\in S_n$ is a reference permutation and has the largest probability value, i.e., the mode of the distribution. The probability of every permutation $\sigma\in S_n$ decays exponentially as its distance $d(\sigma,\sigma_0)$ increases, and the dispersion parameter $\theta$  controls this decay.  The normalization constant $\psi$ can be easily computed for the Kendall's $\tau$ distance (Eq.~\eqref{eq:kendall_decomp}) as well as for the Hamming, Cayley and Ulam distances~\cite{IruCalLoz2016permallows}.

@@

The exact MLE for $\sigma_0$\MANUEL{X MLE for $\sigma_0$ or for $\hat{\sigma}_0$?} is given by the well-known Kemeny ranking~\cite{DwoKumNao2001rank}. Unfortunately,  obtaining such ranking is NP-hard~\cite{DwoKumNao2001rank}. An alternative to the Kemeny ranking is the Borda ranking~\cite{AliMei2011kemeny}, which has several advantages: (i) it requires polynomial computational time; (ii) it  guaranties  high quality parameters as an estimator for the MM~\cite{CarProSha2013votes}; and (iii) in a general optimization perspective, Borda is an approximation to Kemeny~\cite{CopFleRur2010ordering}.\MANUEL{X I'm not sure what the latter means}

@@

\begin{algorithm}[t]
 \caption{UMM: Unbalanced Mallows Model for Black-Box Optimization}
 \label{alg:umm}
  \begin{algorithmic}[1]
    \Require $\minit$: number of initial permutations, $\FEmax$: total budget, $\rho$: learning rate
    %\\ ?, $\FEmax$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
    \State $S \assign $ generate $\minit$ permutations uniformly at random
    \State $F \assign \{ f(\sigma) \mid \sigma \in S\}$ \Comment{Evaluate them}
    \For{$i \assign \minit$ \textbf{to} $\FEmax$ \text{evaluations}}
    \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$\label{line:uborda}
    \State decrease on $\theta$\label{line:theta}
    \State $\sigma \assign$ sampled from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
    \State $S \assign S \cup \{\sigma\}$, $\quad F \assign F \cup \{f(\sigma)\}$
    \EndFor 
    \State \Return $S$, $F$
  %   \Require $n$: ?, $m$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
  %   \For{$i \assign 1$ \textbf{to} $m$ \text{evaluations}}
  %   \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$
  %   \State linear decrease on $\theta$
  %   \State $\sigma \assign$ sample from a $MM(\hat\sigma_0, \theta)$
  %   \State $S \assign S \cup \{\sigma\}$
  %   \State $F \assign F \cup \{f(\sigma)\}$
  %   \EndFor 
  %   \State \Return $S$, $F$
 \end{algorithmic}
\end{algorithm}

As shown in Algorithm~\ref{alg:umm}, UMM  starts with a \emph{small} sample of $\minit$ permutations that are randomly generated (although it would be possible to employ Latin Hyper-Cube sampling or other methods) and evaluated. Then, using the current sample $S$, the uBorda algorithm is applied to find a reference permutation $\hat{\sigma}_0$ (line~\ref{line:uborda}). Together with the parameter $\theta$, they define the Mallows model $MM(\hat\sigma_0, \theta)$. A decreasing $\theta$ parameter (line~\ref{line:theta}) implies a decreasing variance of the Mallows model (Eq.~\ref{eq:MM}), more details in Section~\ref{sec:sampling}. Then, we sample a new candidate permutation from  $MM(\hat\sigma_0, \theta)$ (line~\ref{line:sample}). This candidate permutation (solution) is evaluated and added to the sample $S$. The process is repeated up to a given maximum budget of evaluations $\FEmax$.\MANUEL{We do not explain the initial value of $\theta$ or how we decrease it. Can we give a bit more detail?}

The key idea behind UMM is that the probabilistic Mallows model (MM) used in
UMM does not simply represent the current sample $S$. Instead, the model
assigns higher probability values to the permutations with known (or expected)
minimal fitness values. This is achieved by means of the uBorda ranking
(Section~\ref{sec:uBorda}), which can be seen as replicating each permutation
$\sigma$ in the sample a number of times inversely proportional to its fitness
$f(\sigma)$, i.e., it behaves as if there were many copies of the best
solutions and few of the bad ones. A crucial property of UMM is that the
computational complexity of working with the original sample and with the
transformed sample is the same, i.e., polynomial. The most important steps are
the sampling of a new permutation from $MM(\sigma_0,\theta)$ (line~\ref{line:sample} in Algorithm~\ref{alg:umm}) and the update
(\emph{learning}) of the parameters $\sigma_0$ $\theta$ and $\rho$ (line~\ref{line:uborda} in Algorithm~\ref{alg:umm}).\MANUEL{X Is
  this $\rho$ the same we explained earlier for the weights?}

@@

\subsubsection{Learning step.}\ekhine{ }{conceptuaplemet sampling va después}
\ekhine{As explained above, the parameter $\theta$ is updated deterministically at the
sampling step.}{At each iteration we start by fitting a MM to the sample $S$, which implies estimating $\hat\sigma_0$ and $\theta$. $\theta$ is updated deterministically at the sampling step.} 
The parameter $\sigma_0$, however, is learned from the current
sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate
is controlled by parameter $\rho \in (0,1)$. \ekhine{, which currently is user-defined. Higher values or $\rho$ bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas smaller values make the sampling more uniform, thus slowing down the learning. }{}
In this paper, the value of $\rho$ is set at each iteration and depends on the set of function evaluations $F$. Parameter $\rho$ is set such that the largest $r_1$\% of the mass of the weights is concentrated in the best $r_2$\% of the solutions in $S$, i.e., 
\begin{equation}
\begin{split}
r_1 \sum_{\sigma\in S}\rho^{f(\sigma)} =  \sum_{\sigma\in S'}\rho^{f(\sigma)} \text{where} |S'| = r_2|S| \land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\geq f(\sigma)
\end{split}
\end{equation}

% As shown in Section~\ref{sec:backgroud}, MM can be approximately learnt in two steps (1) estimating the location parameter $\sigma_0$ and (2) computing the scale parameter $\theta$. Since the scale parameter $\theta$ is set in the sampling step,\MANUEL{X How is it set?} we focus on estimating $\sigma_0$.

@@

\subsubsection{Sampling step.}\label{sec:sampling}
Sampling from $MM(\sigma_0,\theta)$  can be done efficiently~\cite{IruCalLoz2016permallows} given the parameters  $\sigma_0$ and $\theta$, as studied by \citet{Collas,IruCalLoz2016permallows,Regenwetter2004}\MANUEL{Which paper is Collas? It is not found.}\MANUEL{You call this Regenwetter, but the author is Doignon? \ekhine{}{he is another coauthor} But the entry shows only 1 author.} where the computational complexities are $O(n\log n)$ and $O(n^2)$.\MANUEL{respectively?}
%
At each iteration of UMM, $\sigma_0=\hat\sigma_0$ is the permutation obtained as a solution in the previous learning step. The scale $\theta$ controls the variance and expected distance of the MM $\Edist$, i.e., the distance of a random permutation $\sigma\sim MM(\sigma_0, \theta)$ to the location parameter $\sigma_0$.
\MANUEL{X I'm not sure what is $D$, distance from where to where?} It is set automatically by the algorithm to control the diversification and intensification trade off. At the first iteration, it is set so that the expected distance $\Edist$ equals  $\binom{n}{4}$ (half of the distance of the uniform distribution), and then the expected distance linearly decrease until the last iteration where the expected distance equals 1. Despite it is not possible to isolate $\theta$ in the $\Edist$ expression, Equation~\eqref{eq:expectation}, its monotonicity allows the use of bisection methods for efficient solution approximations. \MANUEL{X I don't understand this sentence, sorry. You say it does not have a closed-form, but you show a closed-form. Do you mean that the formula below is only an approximation of $\Edist$? What is low-cost here? what is high-quality?}\MANUEL{X What is $k$ in the equation?}
%
\begin{equation}\label{eq:expectation}
\Edist = \frac{n \cdot \exp(-\theta)}{1-\exp(-\theta)} - \sum_{j=1}^{n}  \frac{j \exp(-j \theta)}{1-\exp(-j\theta)}
\end{equation}

A purely explotation approach would always generate $\sigma_0$ in the sampling stage.\MANUEL{X I'm not sure what ``systematically generate $\sigma_0$'' means in this context.} However, the setting of a decreasing variance \MANUEL{X which setting? $\theta$?}  controls the exploration-explotation tradeoff as a function of the number of function evaluations that the algorithm performs.  Similar approaches have been taken in different contexts~\cite{arza2019approaching}.\MANUEL{there is something wrong with this reference\ekhine{}{en mi mendeley esta bien :s que puede ser?} }

>>>>

\usepackage{algorithm}
\usepackage[noend]{algpseudocode}
%\algrenewcommand{\algorithmiccomment}[1]{\hfill\texttt{//} #1}

\usepackage{booktabs} % For formal tables
\usepackage{enumitem}
\usepackage{afterpage,tabularx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage[square,numbers,sort]{natbib}
% Fixes for using natbib and llncs
\makeatletter
% required for natbib to have "References" printed and as section*, not chapter*
\renewcommand\bibsection%
{\section*{\refname\@mkboth{\MakeUppercase{\refname}}{\MakeUppercase{\refname}}}}
\renewcommand\@biblabel[1]{#1.}
\def\bibfont{\small}
\makeatother

@@

Distance in permutation space can be measured in various ways.  Kendall's-$\tau$ counts the number of items ranked in different order in both permutations:
%
\begin{equation}
\begin{split}
d(\sigma _{1},\sigma_{2})=\left\lvert\Bigl\{(i,j):i<j \land \phantom{}\Bigl(\right.&\bigl(\sigma_{1}(i) < \sigma_{1}(j)\land \sigma_{2}(i)>\sigma_{2}(j)\bigr)\lor \phantom{} \\
 &\bigl(\sigma_{1}(i)>\sigma_{1}(j)\land \sigma_{2}(i)<\sigma_{2}(j)\bigr)\Bigr)\Bigr\}\Bigr\rvert\\
 \end{split}
 \label{eq:kendall_decomp}
\end{equation}

An equivalent definition for the Kendall's-$\tau$ distance counts the number of adjacent swaps that have to be made in $\sigma_1^{-1}$ to obtain $\sigma_2^{-1}$ and, therefore, it is sometimes called \emph{swap distance}~\citep{ZaeStoBar2014:ppsn}.

@@

\subsection{Distributions over permutations}
Distributions over permutations~\cite{critchlow91} are functions that assign a probability value  $\Prob(\sigma)\in[0,1]$ to each permutation $\sigma \in S_n$. One of the most popular distributions is the Mallows Model (MM), which is considered as an analogous to the Gaussian distribution for permutations. The MM defines the probability of each permutation $\sigma\in S_n$ as follows:
%
\begin{equation}\label{eq:MM}
\Prob(\sigma \mid \sigma_0, \theta )= \frac{\exp(-\theta d(\sigma, \sigma_0))}{\psi} \sim MM(\sigma_0, \theta)
\end{equation}
%
with two parameters, $\theta$ and $\sigma_0$. Permutation $\sigma_0\in S_n$ is a reference permutation and has the largest probability value, i.e., the mode of the distribution. The probability of every permutation $\sigma\in S_n$ decays exponentially as its distance $d(\sigma,\sigma_0)$ increases, and the dispersion parameter $\theta$  controls this decay.  The normalization constant $\psi$ can be easily computed for the Kendall's $\tau$ distance (Eq.~\eqref{eq:kendall_decomp}) as well as for the Hamming, Cayley and Ulam distances~\cite{IruCalLoz2016permallows}.

@@

The exact MLE for $\sigma_0$ is given by the well-known Kemeny ranking~\cite{DwoKumNao2001rank}. Unfortunately,  obtaining such ranking is NP-hard~\cite{DwoKumNao2001rank}. An alternative to the Kemeny ranking is the Borda ranking~\cite{AliMei2011kemeny}, which has several advantages: (i) it requires polynomial computational time; (ii) it  guaranties  high quality parameters as an estimator for the MM~\cite{CarProSha2013votes}; and (iii) in a general optimization perspective, Borda is an approximation to Kemeny~\cite{CopFleRur2010ordering}.\MANUEL{I'm not sure ``in a general optimization perspective'' means}

@@

\begin{algorithm}[t]
 \caption{UMM: Unbalanced Mallows Model for Black-Box Optimization}
 \label{alg:umm}
  \begin{algorithmic}[1]
    \Require $\minit$: number of initial permutations, $\FEmax$: total budget, $\rho$: learning rate
    %\\ ?, $\FEmax$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
    \State $S \assign $ generate $\minit$ permutations uniformly at random
    \State $F \assign \{ f(\sigma) \mid \sigma \in S\}$ \Comment{Evaluate them}
    \For{$i \assign \minit$ \textbf{to} $\FEmax$ \text{evaluations}}
    \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$\label{line:uborda}
    \State Decrease  $\theta$\label{line:theta}\Comment{Eq.~\ref{eq:expectation:}}
    \State $\sigma \assign$ sampled from  $MM(\hat\sigma_0, \theta)$\label{line:sample}
    \State $S \assign S \cup \{\sigma\}$, $\quad F \assign F \cup \{f(\sigma)\}$
    \EndFor 
    \State \Return $S$, $F$
  %   \Require $n$: ?, $m$: budget of evaluations,\newline $S$: initial permutations, $F$: their fitness values
  %   \For{$i \assign 1$ \textbf{to} $m$ \text{evaluations}}
  %   \State $\hat\sigma_0 \assign \text{uBorda}(S, F,\rho)$
  %   \State linear decrease on $\theta$
  %   \State $\sigma \assign$ sample from a $MM(\hat\sigma_0, \theta)$
  %   \State $S \assign S \cup \{\sigma\}$
  %   \State $F \assign F \cup \{f(\sigma)\}$
  %   \EndFor 
  %   \State \Return $S$, $F$
 \end{algorithmic}
\end{algorithm}

As shown in Algorithm~\ref{alg:umm}, UMM  starts with a \emph{small} sample of $\minit$ permutations that are uniformly randomly generated (although it would be possible to employ Latin Hyper-Cube sampling or other methods) and evaluated. Then, using the current sample $S$, the uBorda algorithm is applied to find a reference permutation $\hat{\sigma}_0$ (line~\ref{line:uborda}). Together with the parameter $\theta$, they define the Mallows model $MM(\hat\sigma_0, \theta)$. A decreasing $\theta$ parameter (line~\ref{line:theta}) implies a decreasing variance of the Mallows model (Eq.~\ref{eq:MM}), as explained in  Section~\ref{sec:sampling} below. Then, we sample a new candidate permutation from  $MM(\hat\sigma_0, \theta)$ (line~\ref{line:sample}). This candidate permutation (solution) is evaluated and added to the sample $S$. The process is repeated up to a given maximum budget of evaluations $\FEmax$.\MANUEL{If the learning step happens later, when does it happen here?}
\MANUEL{We do not explain the initial value of $\theta$ or how we decrease it. Can we give a bit more detail?}

The key idea behind UMM is that the probabilistic Mallows model (MM) used in
UMM does not simply represent the current sample $S$. Instead, the model
assigns higher probability values to the permutations with known (or expected)
minimal fitness values. This is achieved by means of the uBorda ranking
(Section~\ref{sec:uBorda}), which can be seen as replicating each permutation
$\sigma$ in the sample a number of times inversely proportional to its fitness
$f(\sigma)$, i.e., it behaves as if there were many copies of the best
solutions and few of the bad ones. A \manuel{crucial}{} property of UMM is that the
computational complexity of working with the original sample and with the
transformed sample is the same, i.e., polynomial. The most important steps are
the sampling of a new permutation from $MM(\sigma_0,\theta)$ (line~\ref{line:sample} in Algorithm~\ref{alg:umm}) and the update
(\emph{learning}) of the parameters $\sigma_0$, $\theta$ and $\rho$ (line~\ref{line:uborda} in Algorithm~\ref{alg:umm}).

@@

\subsubsection{Learning step.}
At each iteration we start by fitting a MM to the sample $S$, which implies
estimating $\hat\sigma_0$ and $\theta$. Parameter $\theta$ is updated
deterministically at the sampling step.  The parameter $\sigma_0$, however, is
learned from the current sample $S$ by using the uBorda algorithm with a weight
$w(\sigma)=\rho^{f(\sigma)}$ (Section~\ref{sec:uborda}). The learning rate is
controlled by parameter $\rho \in (0,1)$. %
In UMM, the value of $\rho$ depends on the set of true fitness evaluations $F$
and it is set at each iteration such that the largest $r_1$\% of the mass of
the weights is concentrated in the best $r_2$\% of the solutions in $S$, i.e.,
\MANUEL{Let's talk about this Equation tomorrow}
%
\begin{equation}
\begin{split}
r_1 \sum_{\sigma\in S}\rho^{f(\sigma)} =  \sum_{\sigma\in S'}\rho^{f(\sigma)} \quad\text{where}\quad |S'| = r_2|S| \land \forall \sigma'\in S', \sigma\in S\setminus S' : f(\sigma')\geq f(\sigma)
\end{split}
\end{equation}
%
\MANUEL{should the $\sigma$ be the same on both sides of $=$?}
where $S'$ contains the $r_2|S|$ elements of $S$ with the best corresponding fitness values in $F$. The values of $r_1, r_2 \in (0, 1)$ are user-defined and control the balance between exploration and intensification: higher values bias the sampling towards the best solutions found so far, which may lead to quick stagnation, whereas smaller values make the sampling more uniform, thus slowing down the learning.

% As shown in Section~\ref{sec:backgroud}, MM can be approximately learnt in two= steps (1) estimating the location parameter $\sigma_0$ and (2) computing the scale parameter $\theta$. Since the scale parameter $\theta$ is set in the sampling step,\MANUEL{X How is it set?} we focus on estimating $\sigma_0$.

@@

\subsubsection{Sampling step.}\label{sec:sampling}
Sampling from $MM(\sigma_0,\theta)$  can be done efficiently~\cite{IruCalLoz2016permallows} given the parameters  $\sigma_0$ and $\theta$, as studied by \citet{Collas,IruCalLoz2016permallows,DoiPekReg2004rank}\MANUEL{Which paper is Collas? It is not found.} where the computational complexities are $O(n\log n)$ and $O(n^2)$.\MANUEL{respectively?}
%
At each iteration of UMM, $\sigma_0=\hat\sigma_0$ is the permutation obtained in the previous learning step. The scale $\theta$ controls the variance and expected distance $\Edist$ of the MM, i.e., $D$ is the distance of a random permutation $\sigma\sim MM(\sigma_0, \theta)$ to the location parameter $\sigma_0$:\MANUEL{where does this formula come from?}
%
\begin{equation}\label{eq:expectation}
\Edist = \frac{n \cdot \exp(-\theta)}{1-\exp(-\theta)} - \sum_{j=1}^{n}  \frac{j\cdot\exp(-j  \theta)}{1-\exp(-j\theta)}
\end{equation}

Parameter $\theta$ is set automatically by the algorithm to control the diversification and intensification trade off. At the first iteration, $\theta$ is set so that  $\Edist = \binom{n}{4}$ (half of the distance of the uniform distribution), and then the expected distance linearly decreases until the last iteration where $\theta$ is set such that $\Edist = 1$. Despite it is not possible to isolate $\theta$ in Eq.~\eqref{eq:expectation}, its monotonicity allows the use of bisection methods for efficiently finding an approximation.

A purely explotation approach would always generate $\sigma_0$ in the sampling stage. However, the setting of a decreasing variance \MANUEL{But we are controlling the expectation not the variance???}  controls the exploration-explotation tradeoff as a function of the number of function evaluations that the algorithm performs.  Similar approaches have been taken in different contexts~\cite{ArzCebPer2019qap}.
